Follow-up Samples

### 2016-12-07 submission/manuscript_outline_20161024.Rmd
incorporate Fig 6 into manuscript and tab S5

@@

@@

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Many of the OTUs identified classified to normal flora bacterium [Table S4].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.  Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.  That is to say that the bacteria are consistently present even after removal of the lesion by surgery.  Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.

@@

>>>>

# Results of CRC Bacteria
pvalue_adn_crc_summary <- read.csv('../results/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)

@@

***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].  The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].

@@

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Many of the OTUs identified classified to normal flora bacterium [Table S4].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

Naturally, it is curious that normal staples of many screening studies such as Fusobacterium, Parvimonas, and Peptostreptococcus were not present in the majority of the training models.  One potential explanation for this is that FIT provides the same information to the model as these three organisms and so the model uses FIT preferentially over them.  This has been suggested to be the case in a previous study (**insert Baxter Study**).  It is also possible that these specific bacteria play a major role in the progression to carcinoma but may not be as important in the initiation of an adenoma, which would be supported by our data [Figure 5].  Regardless, our study does not argue against the importance of these bacterium in CRC initiation or pathogenesis but rather that the model does not utilize these specific bacteria for prediction purposes.  Another potential reason why we did not identify the "usual suspects"" is that these bacteria may not change much between initial and follow up samples in those with an identified lesion.  That is to say that the bacteria are consistently present even after removal of the lesion by surgery.  Finally, it is likely that within our test set there was not enough indviduals in which detection  was made or relative abundance high enough for these bacteria to be significant using a paired wilcoxson test.

@@

**Figure 6: Previously Associated CRC Bacteria in Initial and Follow up Samples.**  A) Carcinoma initial and follow up samples. There was a significant difference in initial and follow up sample for the OTUs classfied as Peptostreptococcus assacharolytica (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas stomatis (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`).  B) Adenoma initial and follow up samples.  There were no significant differences between initial and follow up.
